---
title: "Survival Theory Course"
subtitle: "Frailty"
author: Andrea Discacciati
output: slidy_presentation
editor_options: 
  chunk_output_type: console
---

``` {r setup, include = FALSE}
library(dplyr)
library(ggplot2)

knitr::opts_chunk$set(fig.width=7, fig.asp=.5, fig.align = 'center') 
```


```{r, eval = FALSE, include = FALSE}
# Outline
1. Univariate frailty models

2. Multivariate (shared) frailty models
```

 


# Frailty and proportional fraily model

A _frailty_ is a latent random variable $Z$ that captures heterogeneity in a population.

In the univariate frailty model (at most one event per individual), the hazard rate of an individual is given as the product of an
individual specific frailty $Z$ and a basic rate $\alpha(t)$, such that 

\begin{align*}
\alpha(t|Z) &= Z\alpha(t)
\end{align*}

Note that $Z$ and $\alpha(t)$ are _not_ observable. What may be observed in a
population is not the individual hazard rates $\alpha(t|Z)$, but the net result for a number
of individuals with differing frailties, or what we term the _population hazard
rate_.





# Population survival

Given $Z$, (conditional) survival to time $t$ is given by

\begin{align*}
S(t|Z) &= \exp\left(-Z\int_0^t \alpha(u) du\right) \\
&= \exp(-Z A(t))
\end{align*}

where $A(t)$ is the cumulative hazard. <!-- As from earlier in the course, $A(t)$ plays an important role in the mathematical development. -->

The _population survival function_ is found by integrating $S(t|Z)$ over the distribution
of $Z$, such that

\begin{align*}
S(t) &= \int_0^\infty f_Z(z) \exp(-zA(t)) dz \\ &= E_Z\left(\exp(-ZA(t))\right)
\end{align*}

$S(t)$ has a population averaged interpretation. If there are no covariates, $S(t)$
may be seen as a weighted average of individual survival curves, where the weighing
depends on the distribution of $Z$. 


# Population hazard

It is useful to introduce the _Laplace transform_ of $Z$, defined by

\begin{align*}
\mathcal{L}(c) &= E_Z\left(\exp(-cZ)\right)  \\ &= \int_0^{\infty} f_Z(z) \exp(-cz) dz
\end{align*}

and to write

\begin{align*}
S(t) &= \mathcal{L}(A(t))
\end{align*}

The _population hazard rate_, denoted by $\mu(t)$, may be found by differentiating $-\log(S(t))$, so that

\begin{align*}
\mu(t) &= \alpha(t)\frac{-\mathcal{L}'(A(t))}{\mathcal{L}(A(t))} \\
& = \alpha(t) E(Z|T>t)
\end{align*}

$\mu(t)$ may be seen as a
weighted average of individual hazards of individuals alive at time $t$, where the weighing
depends on the distribution of $Z$ among the individuals alive at time $t$.

The difference between the conditional and population hazard is determined by $\frac{-\mathcal{L}'(A(t))}{\mathcal{L}(A(t))} = E(Z|T>t)$,
which is a decreasing function of $t$.


# Notes

-  The proportional frailty model is chosen for mathematical convenience. It
represents a specific simplified view of how individual heterogeneity might act.
There is no reason why heterogeneity should be determined at time zero, that it
should be constant over time, or that it should act in a proportional manner.

-  Nevertheless, this model is _useful_ to explain the behavior of the hazard (selection of healthier subjects over time) 
and of the hazard ratio (deviations from proportionality of the hazards).

- Given the results presented in the previous slides, it's natural to use frailty distributions with
a tractable and explicit Laplace transform. Today, we will focus on the _gamma distribution_.
See ABG Chapter 6 for other distributions.



# Gamma frailty distribution

A common distributional assummption for $Z$ is the _gamma distribution_, which has
    probability density function

\begin{align*}
f_Z(z) &= \frac{\nu^\eta}{\Gamma(\eta)z^{\eta-1}\exp(\nu z)}
\end{align*}

where $\Gamma(\eta)=\displaystyle\int_0^\infty
\displaystyle\frac{1}{z^{1-\eta}\exp(z)}dz$, $\nu$ is a scale parameter, and
$\eta$ is a shape parameter.

Given this parametrisation $E(Z)=\eta/\nu$ and $Var(Z)=\eta/\nu^2$.

For identifiability, we let the frailty distribution $Z$ have mean equal to 1, so
that $\eta=\nu$, which has $Var(Z)=\nu^{-1}$. 

We now parameterise the distribution using the variance $\delta=\nu^{-1}$.
$\delta$ is a natural measure of the degree of heterogeneity of the population.



```{r, echo = FALSE}
expand.grid(z = seq(0,4,length=501),
            delta = c(0.01, 0.4,1,5)) %>%
    mutate(y=stats::dgamma(z, shape=1/delta, scale=delta),
           delta=factor(delta)) %>%
    ggplot(aes(x=z, y=y, col=delta)) + geom_line(lwd = 2) +
    ylim(0, 4) + xlab("z") +
    ylab("Probability density") + theme_bw(base_size = 14)
```


# Gamma frailty distribution

The Laplace transform is then given as

\begin{align*}
\mathcal{L}(c) &= (1+\delta c)^{-1/\delta}
\end{align*}

We can now express the population survival as

\begin{align*}
S(t) &= \mathcal{L}(A(t)) = (1+\delta A(t))^{-1/\delta}
\end{align*}

and the population hazard rate as

\begin{align*}
\mu(t) &= -\alpha(t) \frac{\mathcal{L}'(A(t))}{\mathcal{L}(A(t))} = \frac{\alpha(t)}{1+\delta A(t)}
\end{align*}

When $\delta \downarrow 0$, the variability of the frailty distribution tends to 0 and $\mu(t) \approx \alpha(t)$. As the variance $\delta$ increases,
the population hazard function assumes the frailty shape of a hazard function that is "dragged down" (see next slide).


# Gamma frailty distribution: population hazard rate

Population hazard rate $\mu(t)$ when $\alpha(t) = 1$ and $\alpha(t) = t^2$, for different
values of $\delta$.



``` {r, echo = FALSE, warning =FALSE}
expand.grid(t = seq(0,6,length=51),
            delta = c(0, 0.4,1,2,5),
            alpha = c("1", "t^2")) %>%
    mutate(A = ifelse(alpha == "1", t, t^3/3),
           mu = ifelse(alpha == "1", 1, t^2)/(1+delta*A),
           delta=factor(delta)) %>% 
    ggplot(aes(x=t, y=mu, col=delta)) + geom_line(lwd = 2) +
    ylim(0,2.5) + xlab("Time") + facet_grid(~ alpha, labeller = label_both) + 
    ylab("Population hazard rate") + theme_bw(base_size = 14)
```


When $\delta =  0$ there is no frailty effect and $\mu(t) = \alpha(t)$.

The presence of frailty ($\delta>0$) pulls the population hazard rate down, transforming even an 
ever-increasing basic rate ($\alpha(t) =t^2$, right panel) into a quite different shape.

This can be explained by "frail" individuals (high levels of $Z$) developing the event earlier, leaving a 
population consisting of "less frail" individuals (low levels of $Z$). Intuitively, does 
this agree with the fact that $\mu(t) = \alpha(t) E(Z|T>t)$?

If frailty is present, the observed population hazard rate does not say anything about the development at the 
individual level, but it is mixed up with selection effects.


# Gamma frailty distribution: effect on hazard ratio

Assume that the basic hazard rates in two risk groups are $\alpha(t)$ and $r \alpha(t)$ with $r>1$.

Assume that the frailty variables have the same gamma distribution with variance $\delta$ in both groups (ie, we 
assume proportionality of the hazards to hold at the individual level).

The population hazard ratio is

\begin{align*}
\frac{\mu_2(t)}{\mu_1(t)} &=r \frac{1+\delta A(t)}{1+r\delta A(t)}
\end{align*}

This expression is decreasing in $t$ and if $A(t) \rightarrow \infty$ when $t$ increases then the ratio approaches 1.

Typically, if we assume proportionality at the level of the individual hazards, that proportionality
will disappear at the population level.

``` {r, echo = FALSE, warning =FALSE}
expand.grid(t = seq(0,2,length=51),
            delta = c(0, 0.5, 2),
            alpha = c("1", "t^2"), 
            r = 4) %>%
    mutate(A = ifelse(alpha == "1", t, t^3/3),
           phr = r * (1+delta*A) / (1+r*delta*A),
           delta=factor(delta)) %>% 
    ggplot(aes(x=t, y=phr, col=delta))+ geom_line(lwd = 2) +
    xlab("Time") + facet_grid(r ~ alpha, labeller = label_both) +
  ylim(0,4) +
    ylab("Population hazard ratio") + theme_bw(base_size = 14)
```

In this case, the observed hazard ratio will be decreasing due to frailty but $\frac{\mu_2(t)}{\mu_1(t)}>1$. With other
frailty distributions it may cross below one. With the stable frailty distribution, on the other hand, proportionality of the hazards holds
at the individual and at the population level (but the proportionality constants are different); see ABG 6.5.



<!-- # Frailty and estimation: identifiability -->

<!-- When there is at most one event for each individual, frailty models will often not be _identifiable_.  -->
<!-- In general, different frailty models can result in a given population hazard rate. An exception is when  -->
<!-- it is possible to hypothesise a parametric model class for the basic hazard rate $\alpha(t)$. -->

<!-- This may not be enough, though. With gamma frailty, we can divide and multiply $Z$ and $\alpha(t)$ by the same constant $c$,  -->
<!-- thereby staying in the gamma model, but with a new scale parameter $\alpha = c / \nu$ and the same shape parameter $\eta$. That's  -->
<!-- the reason for the restriction $\eta = \nu$. -->

<!-- See ABG 6.4.1 for a parametric unit-mean-Gamma--Weibull model and ABG 6.9 for a  -->
<!--   parametric frailty model with Weibull basic rate . -->

<!--  This identifiability problem can be generally avoided in the multivariate case (clustered or recurrent events). -->


# Identifiability

When there is at most one event for each individual, frailty models will often not be _identifiable_. 

Observing a population hazard ratio like the ones we saw earlier is compatible with two explanations:

1. The proportional hazard assumption holds in the conditional model ($\alpha_2(t|Z) = Zr\alpha_1(t)$). What we observe is a result of unobserved heterogeneity (frailty).

2. There is no heterogeneity. What we observe is the time-dependent effect of the covariate ($\alpha_2(t) = r(t) \alpha_1(t)$).

Based on the (observed) population hazard rates alone, it is impossible to decide whether frailty is present or not. To fit a frailty model, we would first need to assume a conditional proportional hazards assumption and then select 
a parametric model class for the basic hazard rate $\alpha(t)$.

<!-- This may not be enough, though. With gamma frailty, we can divide and multiply $Z$ and $\alpha(t)$ by the same constant $c$,  -->
<!-- thereby staying in the gamma model, but with a new scale parameter $c / \nu$ and the same shape parameter $\eta$. That's  -->
<!-- the reason for the restriction $\eta = \nu$. The stable frailty distribution is never identifiable with univariate survival data. -->

<!-- This issue is of course present also with no covariates (see exercises). -->

This identifiability problem can be generally avoided in the multivariate case (clustered or recurrent events).


<!-- # Gamma--Weibull model -->

<!-- In the univariate case, the population hazard function obtained from assuming a Gamma-distributed frailty -->

<!-- \begin{align*} -->
<!-- \mu(t) = \frac{\alpha(t)}{1+\delta A(t)} -->
<!-- \end{align*} -->

<!-- defines a probability distribution $T$.  -->

<!-- If we let the basic hazard be $\alpha(t) = bt^{k-1}$ (Weibull), then $T \sim \textsf{Burr}(\delta, b, k)$ with pdf -->

<!-- \[ -->
<!-- f(t) = bt^{k-1}\left(1 + \frac{\delta b}{k} t^k \right)^{-1-\frac{1}{\delta}} -->
<!-- \] -->

<!-- With covariates, $b$ is replaced by $\exp(\beta^\top x_i)$. -->

<!-- $Var(T)$ increases as $\delta$ increases. As we said, $\delta$ is a measure -->
<!-- of the degree of the heterogeneity in the population. -->

``` {r, eval = FALSE, echo = FALSE, warning =FALSE}
varT <- function(delta, b, k) {
  1/delta * (k/(b*delta))^(2/k) * (beta(2/k+1, 1/delta-2/k) - 1/delta*beta(1/k+1, 1/delta-1/k)^2)
}

expand.grid(delta = seq(0, 1, length = 51)[-1],
            b = 1,
            k = 1:4) %>% 
  mutate(alpha = paste0("t^", k-1),
         var = varT(delta, b, k)) %>% 
  ggplot(aes(x = delta, y = var, color = alpha)) +
 geom_line(lwd = 2) +
    ylim(0, 3) + xlab("delta") +  
    ylab("Var(T)") + theme_bw(base_size = 14)
```



# Multivariate frailty models 

We now consider grouped outcomes, where the groups can be defined in
    terms of
    
  1.  _Clusters_, when several units that may fail are collected in a cluster;
    
  2.  _Recurrent events_, when several successive events of the same type are registered for each individual.
    
For grouped outcomes, we can consider _shared frailty
    models_, ie the same frailty is shared by all the survival time of an 
    individual or cluster $i$.
  
Frailty measures the specific risk level for a cluster or an individual,
and given $Z$ the survival times are independent. We assume the proportional frailty model $\alpha(t|Z_i)=Z_i\alpha(t)$.
    

# Notation    
    
Consider $m$ independent clusters or individuals with index $i$. Let the actual event time be $T_{ij}$ and let the
    observed (possibly right-censored) times be $\tilde{T}_{ij}$  for the $j=1,\ldots,n_i$ cluster units or inter-event times.

The basic rate $\alpha(t)$ is a fixed function and assumed to be the
    same for all $i$, but each cluster or individual has a separate independent
    frailty $Z_i$ with $\boldsymbol{Z}=\{Z_1, \ldots, Z_i, \ldots, Z_m\}$. All of the frailty variables
    are identically distributed.

Lastly, let $D_{ij}$ be binary variables that are 1 if the event time is
    uncensored and are 0 if censored. Let $D_{i\bullet}=\sum_j D_{ij}$,
    which is the number of events for cluster or individual $i$.

# Conditional likelihood

Assume that, conditional on $\boldsymbol{Z}=\boldsymbol{z}$, censoring is independent and noninformative of $\boldsymbol{z}$ (ABGK Assumption IX.3.1 and IX.3.2).

For each cluster of individual $i$, $\tilde{T}_{ij}$ and $D_{ij}$ are combined into the history $H_i=(\tilde{T}_{i1}, D_{i1}, \ldots, \tilde{T}_{in_i}, D_{in_i})$. Let $P(H_i)$
    denote the joint distribution of $H_i$. Conditional with
    respect to $Z_i$, the event times are independent with intensity
    $Z_i\alpha(t)$.
    
ABG argue that their conditional likelihood $P(H_i|Z_i)$ can be deduced (sic) from 

\begin{align}
L_i(\theta) = \alpha(\tilde{T}_i;\theta)^{D_i}\exp\left\{- \int_0^{\tilde{T}_i} \alpha(t;\theta)dt\right\}
\end{align}

as

\begin{align*}
P(H_i|Z_i) &= \prod_{j=1}^{n_i} \left(\left(Z_i\alpha(\tilde{T}_{ij})\right)^{D_{ij}}\exp(-Z_i A(\tilde{T}_{ij}))\right)
\end{align*}

where $\theta$ is dropped from $\alpha$ to simplify notation.


# Unconditional and total likelihood

The unconditional likelihood follows by taking the expectation with respect to frailty variable $Z$

\begin{align*}
P(H_i) &= E_{Z_i}(P(H_i|Z_i))  \\ &= \prod_{j=1}^{n_i} \left(\alpha(\tilde{T}_{ij})^{D_{ij}}\right)
E_{Z_i}\left(Z_i^{D_{i\bullet}}\exp(-Z_i V_i)\right)
\end{align*}

where $V_i=\sum_j A(\tilde{T}_{ij})$.


Applying the fact that the $r\text{th}$ order derivative of a Laplace
    transform is $\mathcal{L}^{(r)}(c)=(-1)^rE(Z^r\exp(-cZ))$, we can
    now write the contribtion of individual or cluster $i$ as

\begin{align*}
P(H_i) &= \prod_{j=1}^{n_i} \left(\alpha(\tilde{T}_{ij})^{D_{ij}}\right) 
(-1)^{D_{i\bullet}}\mathcal{L}^{(D_{i\bullet})}(V_i)
\end{align*}

and the total log-likelihood of $m$ conditionally independent individuals or clusters as

\begin{align*}
\log L &= \sum_{i=1}^m\left( \sum_{j=1}^{n_i} D_{ij}\log\left(\alpha(\tilde{T}_{ij})\right) +
\log\left((-1)^{D_{i\bullet}}\mathcal{L}^{(D_{i\bullet})}(V_i)\right) \right)
\end{align*}

<!-- Estimation comes from maximising over a set of parameters, with -->
<!--     covariance estimation by inversion of the negative Hessian. -->
    
Predicted frailty for cluster or individual $i$ is  $\hat{Z_i}=E(Z_i|H_i) = -\frac{\mathcal{L}^{(D_{i\bullet}+1)}(V_i)}{\mathcal{L}^{(D_{i\bullet})}(V_i)}$  (see ABG 7.2.3).


# Frailty and counting processes

Consider a multivariate counting process $\boldsymbol N = (N_i(t), i=1, \ldots, m)$,
where each $N_i$ can have more than one jump, with individual intensity processes

\[
\lambda^*_i(t)=Z_iY_i(t)\alpha(t)
\]

$\boldsymbol Y=(Y_i(t), i=1, \ldots, m)$ are observable predictable processes
such that, in case of _recurrent events_, $Y_i(t) = 1$  as long as the process is 
observed and 0 afterwards, and, in case of _clustered data_,  $Y_i(t)$ represents
the number of units still at risk. 

<!-- The variables $Z_i$ are always i.i.d. frailties. -->

<!-- <span style="color:grey">Typo in ABG 7.3: $\alpha_i(t)$ should read $\lambda_i(t)$.</span> -->

The complete but unobservable data consists of the triplet $\{\boldsymbol Z, \boldsymbol N, \boldsymbol Y \}$,
while the incomplete but observable data consists of $\{\boldsymbol N, \boldsymbol Y \}$.

The intensity processes $\lambda^*_i(t)$ are relative to the history 
$\mathcal{G}_{t} = \sigma\left\{\boldsymbol Z, \boldsymbol N(u), \boldsymbol Y(u+): 0 \le u \le t \right\}$ partly generated
by unobserved quantities. When calculating likelihood functions, we have to consider
intensities wrt the smaller history generated by the observed data:
$\mathcal{F}_{t} = \sigma\left\{\boldsymbol N(u), \boldsymbol Y(u+): 0 \le u \le t \right\}$, with $\mathcal{F}_{t} \subset \mathcal{G}_{t}$.

# Observed intensity processes

The observed intensity process $\lambda_i(t)$ for the individual $N_i(t)$ is given 
by the innovation theorem

\begin{align}
\lambda_i(t) &= E(E(dN_i(t)|\mathcal{G}_{t^-})|\mathcal{F}_{t^-}) \\ 
&= E(\lambda^*_i(t)|\mathcal{F}_{t^-}) \\
&= Y_i(t)\alpha(t)E(Z_i | \mathcal{F}_{t^-})
\end{align}

The observed intensity process can be written as (ABG 7.3)

\[
\lambda_i(t) = Y_i(t)\alpha(t) \frac{-\mathcal{L}^{(N_i(t^-)+1)}\left(\int_0^{t^-}Y_i(s)\alpha(s)ds\right)}{\mathcal{L}^{(N_i(t^-))}\left(\int_0^{t^-}Y_i(s)\alpha(s)ds\right)}
\]

If the $Z_i$ are unit-mean--Gamma distributed then

\[
\lambda_i(t) = Y_i(t)\alpha(t) \frac{1+\delta N_i(t^-)}{1+\delta\int_0^{t^-}Y_i(s)\alpha(s)ds}
\]

# Likelihood in counting process form

If we specify a parametric class for $\alpha(t)$, we can plug $\lambda_i(t)$ into
the log-likelihood

\[
\log L = \sum_{i=1}^m \int_0^{\tau} \log\lambda_i(s)dN_i(s)-\int_0^{\tau} \lambda_{\bullet}(s;) ds 
\]

Estimates may be found by maximising the likelihood directly and the 
same counting process tools we saw last time can be used for
proving large sample properties of the maximum likelihood estimators.

If, conditional on $\boldsymbol{Z}=\boldsymbol{z}$, censoring is also noninformative on 
$\theta$, then the log-likelihood above is a full marginal log-likelihood for $\alpha(t)$. Otherwise, it's
a partial marginal log-likelihood. 

In other words, the log-likelihood above is either the full or partial log-likelihood for $\theta$ based on 
$\{\boldsymbol N$, $\boldsymbol Y\}$, ie in the marginal/incomplete data experiment when $\boldsymbol Z$ 
is not observed.

Further details are in ABGK IX.2, IX.3, IX.4.1.

<!-- <span style="color:grey">Typo in ABGK Equation (9.3.5): the second product is over $t$, and not over $i$.</span> -->


# Software implementation

Fully parametric frailty models are available here:

- R: `parfm`, `frailtyEM`, `frailtypack`, `rstpm2`

- Stata: `streg`, `merlin`

- SAS: `proc nlmixed`



# Example, Gamma-Weibull model

Using the `parfm` package.

```{r, message = FALSE}
library(parfm)

bc <- read.csv("/Users/anddis/OneDrive - KI.SE/other/teaching/tsa_2022/bc.csv")
bc$id <- 1:nrow(bc)
head(bc)

frailty.mod <- parfm(Surv(t, dead) ~ 1, 
                     dist = "weibull", 
                     frailty = "gamma", 
                     cluster = "id",
                     data = bc)
frailty.mod
```


`Z.hat`, $\hat{Z_i}= -\frac{\mathcal{L'}(A(t))}{\mathcal{L}(A(t))}$

```{r}
Z.hat <- predict(frailty.mod)
hist(Z.hat)
```

```{r, include = FALSE}
hWeibull <- function(t, lambda, rho) lambda * rho * t^(rho-1)
HWeibull <- function(t, lambda, rho) lambda * t^rho
```

`cond.haz`, $\alpha(t| Z) = Z \alpha(t)$

```{r}
timeseq <- seq(0, max(bc$t), length = 501)
alpha.hat <- hWeibull(timeseq, 
                      frailty.mod["lambda", 1], 
                      frailty.mod["rho", 1])
cond.haz <- outer(alpha.hat, Z.hat)
```

`pop.haz`, $\mu(t) = \frac{\alpha(t)}{1+\theta A(t)}$

```{r}
mu.hat <- function(t, lambda, rho, theta) {
  (hWeibull(t, lambda, rho)) / 
    (1 + frailty.mod["theta", 1] * HWeibull(t, lambda, rho))
}

pop.haz <- mu.hat(timeseq, 
                  frailty.mod["lambda", 1], 
                  frailty.mod["rho", 1], 
                  frailty.mod["theta", 1])
```

`pop.haz.empirical`, $\mu(t) = \alpha(t) \bar Z_{T>t}$

```{r}
pop.haz.empirical <- sapply(
  timeseq, 
  function(x) mean(Z.hat[bc$id[bc$t > x]])
) *  alpha.hat
```

```{r, fig.width=7, fig.asp=.7}
matplot(timeseq, cond.haz, type = "l", col = "grey75", lty = 1,
        ylab = "Conditional / population hazard", xlab = "Time",
        main = "Gamma-Weibull frailty model")
lines(timeseq, pop.haz, col = "red", lwd = 5)
lines(timeseq, pop.haz.empirical, type = "s", lwd = 2)
```



